Supporting Information

### 2017-06-15 code/rmarkdown/supporting-information.Rmd
update ms

# Appendix S1
The reliable formulation explicitly considers the probability that planning units are inhabited by features. As a consequence, it may deliver prioritizations that will sufficiently represent an attribute space even if the features do not inhabit several of the planning units when the prioritization is implemented. This behavior is achieved by siting back-up planning units near selected planning units with low occupancy probabilities in the attribute space(s). To ensure that prioritizations are robust against multiple planning units being uninhabited, the problem assigns planning units at multiple backup levels.

Backup levels levels are defined as $r$-levels [similar to failure levels in @r18]. The first backup $r$-level is used to calculate the level of representation when all of the selected planning units are occupied by all $f \in F$. For this scenario, the closest selected planning unit to each demand point $i$ for attribute space $s$ is assigned at $r$-level$=0$. This scenario essentially represents $Y_{fsij}$ in the unreliable formulation. The second backup $r$-level is used to assess the level of representation when the closest planning unit to each demand point $i$ is unoccupied. For this scenario, the second closest planning units are assigned at $r$-level$=1$. The third backup $r$-level is used to assess representation when the first two closest planning units are unoccupied. The third closest planning units are assigned at $r$-level$=2$. Continuing on, in this manner, the selected planning units in a prioritization are assigned to each demand point $i \in I$, attribute space $s \in S$, and each feature $f \in F$ at an $r$-level.

A final backup $r$-level when $r=R$ is used to assess the level of representation when the features $f \in F$ do not occupy any selected planning units in a prioritisation. Each demand point $i \in I$ for each $s \in S$ and $f \in F$ is assigned to an "imaginary" planning unit $j=J$ at $r=R$. The distance variables associated with this imaginary planning unit $d_{fsiJ}$ denote the loss of biological value associated with failing to secure a representative sample of feature $f$ in attribute space $s$. However, the $d$ variables are in distance units which effectively meaningless in this context. Thus these variables are calculated using a failure multiplier ($M$) and the maximum distance between the planning units and the demand points for $f \in F$, $s \in S$ (eqn 5).

@@

Conservation planning problems often use more than several hundred planning units. It is currently not be feasible to solve this problem when considering all possible failure scenarios within a reasonable amount of time (see Appendix S2). As a consequence, the $R$ variable can be any $1 \leq R \leq J-1$. For instance, when $R=3$ only 2 backup levels are considered in addition to the final backup level. Cui _et al._ [-@r16] found that $R=5$ yields similar solutions to $R >> 5$. However, in most cases the decision maker will likely be limited to $R=1$ to obtain prioritizations in a feasible amount of time.

@@

\begin{align*}
R &= \parbox{25em}{number of failure levels} \tag*{eqn 6a} \\
%
M &= \parbox{25em}{failure multiplier} \tag*{eqn 6b}\\
%
\end{align*}

@@

\begin{align*}
Y_{fsijr} &= \begin{cases}
    1, & \parbox{25em}{if demand point $i$ is assigned to planning unit $j$ for feature $f$ in space $s$ at back-up level $r$. } \tag*{eqn 7a} \\
    0, & \parbox{25em}{otherwise} \\
  \end{cases} \\
%
P_{fsijr} &= \parbox{25em}{probability that demand point $i$ is assigned to planning unit $j$ at back-up level $r$ for feature $f$ and space $s$} \tag*{eqn 7b}\\
%
\end{align*}

@@

The reliable formulation is non-linear. However, the non-linear components can be linearized. First--as discussed in the main text--the expression $X_j X_k$ in (eqn 3a) can be linearized using methods described by Beyer et al. [-@r426]. Second, the expression $P_{fsijr} Y_{fsijr}$ in (eqn 8a) can be linearized using techniques described by Sherali and Alameddine [-@r20] as implemented in Cui et al. [-@r16].

@@

We simulated conservation planning data sets containing different numbers of features and planning units (see Figure S2 for amounts). For a given number of planning units and features, a data set was simulated using the following steps. First, the planning units were simulated as squares that were arranged in a grid. Second, the planning unit costs were set to one. Third, the spatial distribution of each feature in the planning units were simulated using binary, spatially autocorrelated random fields [using the \texttt{RandomFields R} package; @r494; @r493]. Fourth, if a feature was found to occupy less than two planning units, the feature was set to occupy two randomly selected planning units. All occupied planning units were set to have a probability of occupancy of `r benchmark_parameters$occupancy_probability` for all features. Fifth, similar to the simulated data sets described in the main text, a geographic attribute space was generated for each feature. The features' demand points were set as the centroids of the planning units that they occupied, and were weighted by the units' probability of occupancy. After simulating the data sets, solutions were generated using the \texttt{raptr R} package and Gurobi (version 7.0.2).

We generated solutions for the conservation planning data sets using different formulations and parameters. Specifically, the solutions were generated using the unreliable and reliable formulation, and several boundary length multipliers (`r paste(unique(benchmark_parameters$blm), sep = "and")`). The solutions were generated using amount-based targets of `r benchmark_parameters$amount_target*100` % and space-based targets of `r benchmark_parameters$space_target*100` %. To assess variation in solve times, we generated and solved `r benchmark_parameters$replicates` random data sets per combination of number of features and planning units. All problems were solved to within `r general_parameters$MIPGap * 100` % of optimality and were solved using a single thread on a system with 64 GB RAM and an Intel(R) Xeon (R) CPU (E5-2630 v3; 2.40 GHz).

@@

The benchmark analysis revealed that moderately-sized conservation problems can be solved in a relatively short period of time (Figure S2). Problems were solved much quicker when using the unreliable formulation than when using the reliable formulation. The solve times could be further reduced by using more threads. We were not able to determine how long problems bigger problems would take to solve using the reliable formulation because the system did not have enough RAM. Future research could involve implementing more advanced methods to solve the optimization problems [eg. as described in @r16; @r15].

@@

# Figures
```{r, echo = FALSE, fig.height = 8, fig.width = 8.5, fig.cap = "Attribute spaces used in the first case-study. Each panel shows a the distribution of a solution in environmental space and how it samples the realized niche for a different species. The left column of panels shows the solution generated using amount targets. The right column of panels shows the solution generated using amount and space targets. Each column of panels corresponds to a different species. Hexagons show the distribution of demand points. The color of each hexagon denotes the weighted frequency of demand points inside it. Points represent the environmental conditions associated with planning units inside the species geographic range that were selected for preservation in a given solution."}
## prepare data
cs1_pu_niche_data <- plyr::ddply(cs1_spp_results,
  c("Species", "Prioritisation"),
  function(x) {
    # init
    curr_prioritisation <- cs1_prioritisations[[as.character(x$Prioritisation)]]
    curr_spp_pos <- match(x$Species[1], curr_prioritisation@data@species[[1]])
    curr_ids <- curr_prioritisation %>%
                slot("data") %>%
                slot("attribute.spaces") %>%
                `[[`(1) %>%
                slot("spaces") %>%
                `[[`(curr_spp_pos) %>%
                slot("planning.unit.points") %>%
                slot("ids")
    curr_sel_pos <- which(curr_ids %in% which(as.logical(
                          raptr::selections(curr_prioritisation))))
    # extract planning units
    curr_data <- curr_prioritisation %>%
                slot("data") %>%
                slot("attribute.spaces") %>%
                `[[`(1) %>%
                slot("spaces") %>%
                `[[`(curr_spp_pos) %>%
                slot("planning.unit.points") %>%
                slot("coords") %>%
                as.data.frame() %>%
                magrittr::set_names(c("Niche1", "Niche2")) %>%
                dplyr::mutate(Species = x[["Species"]][[1]],
                              Prioritisation = x[["Prioritisation"]][[1]],
                              Solution <- "discarded")
    curr_data$Solution[curr_sel_pos] <- "selected"
    # return data
    return(curr_data)
})
cs1_pu_niche_data %<>% dplyr::filter(Solution == "selected")

@@

```{r, echo = FALSE, message = FALSE, fig.height = 5.0, fig.width = 4.5, fig.cap = "Time required to solve different sized problems using different parameters and formulations. Points represent average times and bars show standard errors. Time is shown on a log$_{10}$ scale."}
# prepare data
plot_data <- benchmark_results %>%
  dplyr::mutate(blm = paste0("BLM = ", blm)) %>%
  dplyr::mutate(number_features = factor(number_features)) %>%
  dplyr::mutate(formulation = paste0(formulation, " formulation")) %>%
  dplyr::mutate(number_planning_units = factor(number_planning_units))
# make plot
ggplot2::ggplot(plot_data,
                ggplot2::aes(x = number_planning_units, y = time,
                             color = number_features)) +
  ggplot2::stat_summary(fun.y = mean, geom = "point",
                        position = ggplot2::position_dodge(0.95)) +
  ggplot2::stat_summary(fun.data = ggplot2::mean_se, geom = "errorbar",
                        position = ggplot2::position_dodge(0.95)) +
  ggplot2::facet_grid(blm ~ formulation ) +
  ggplot2::scale_y_continuous(trans = "log10") +
  ggplot2::xlab("Number of planning units") +
  ggplot2::ylab("Solve time (s)") +
  ggplot2::labs(color = "Number of features") +
  ggplot2::theme(
    axis.line.x = ggplot2::element_line(),
    axis.line.y = ggplot2::element_line(),
    strip.background = ggplot2::element_rect(fill = "grey20"),
    strip.text = ggplot2::element_text(color = "white"),
    legend.position = "bottom")
```

@@

```{r, echo = FALSE, results = "asis"}
# make table
terms <- c(
  "$F$", "set of biodiversity features (indexed by $f$)",
  "$J$", "set of planning units (indexed by $j$)",
  "$q_{fj}$", "probability of feature $f$ occupying planning unit $j$",
  "$e_{jk}$", "shared edge between planning unit $j$ and planning unit $k$. Note that when $j==k$ this used to parametrize exposed edges with no neighbours.",
  "$S$", "set of attribute spaces (indexed by $S$)",
  "$I_{fsi}$", "set of demand points (indexed by $i$) for a feature $f$ in attribute space $s$",
  "$\\lambda_{fsi}$", "set of weights for demand point $i$ for feature $f$ in attribute space $s$",
  "$d_{fsij}$", "distance between demand point $i$ and planning unit $j$ for feature $f$ in attribute space $s$",
  "$\\delta_{fsi}$", "the distance between each demand point $i$ and the centroid of the demand points $I$ for feature $f$ in attribute space $s$",
  "$T_f$", "amount target for feature $f$",
  "$\\tau_{fs}$", "space-based target for feature $f$ in attribute space $s$",
  "$X_j$", "binary decision variable controlling if a planning unit is selected for preservation (1) or discarded (0)",
  "$Y_{fsij}$", "binary decision variable indicating if planning unit $j$ is assigned to demand point $i$ for species $s$ in a attribute space $s$ when determining the amount of the attribute space sampled by the selected planning units"
)
term_data <- data.frame(
  Symbol = terms[seq(1, length(terms), 2)],
  Description=terms[seq(2, length(terms), 2)]
)
term_data$Description <- paste0("\\parbox[t][][t]{15cm}{",
                                term_data$Description, "}")

# render table
curr_file <- tempfile(fileext = ".tex")
tmp <- Hmisc::latex(term_data,
                    file = curr_file,
                    digits = 2, rowname = NULL, first.hline.double = FALSE,
                    booktabs = TRUE, here = TRUE,
                    colheads = c("Symbol", "Description"),
                    col.just = c("c", "l"),
                    caption = paste0("Symbols and descriptions of terms used ",
                                     "in the formulation of the unreliable ",
                                     "representative and adequate ",
                                     "prioritization (URAP) problem."))
curr_file %>%
  readLines() %>%
  `[`(-1) %>%
  paste0(collapse = "\n") %>%
  gsub(pattern = "toprule", replacement = "toprule[1pt]", fixed = TRUE) %>%
  gsub(pattern = "bottomrule", replacement = "bottomrule[1pt]",
       fixed = TRUE) %>%
  cat()
```

>>>>

# Appendix S1

The reliable formulation explicitly considers the probability that planning units are inhabited by features. As a consequence, it may deliver prioritizations that sufficiently conserve an attribute space even if the features do not inhabit several of the planning units when the prioritization is implemented. This behavior is achieved by siting back-up planning units near selected planning units with low occupancy probabilities in the attribute space(s). To ensure that the solution is robust against multiple planning units being uninhabited, demand points are treated as being conserved by different planning units at different backup levels.

Backup levels are defined as $r$-levels [similar to failure levels in @r18]. The first backup $r$-level is used to calculate how well the demand points are conserved assuming that all the selected planning units are occupied by all $f \in F$. For this scenario, the closest selected planning unit to each demand point $i$ for attribute space $s$ is assigned at $r$-level = 0. This scenario essentially serves as $Y_{fsij}$ in the unreliable formulation. The second backup $r$-level is used to assess how well each demand point is conserved when the closest planning unit to each demand point $i$ is unoccupied. For this scenario, the second closest planning units are assigned at $r$-level = 1. The third backup $r$-level is used to assess how well demand points are captured when the first two closest planning units are unoccupied. The third closest planning units are assigned at $r$-level = 2 . Continuing on, in this manner, the selected planning units in a prioritization are assigned to each demand point $i \in I$, attribute space $s \in S$, and each feature $f \in F$ at an $r$-level.

A final backup $r$-level ($r=R$) is used to describe how well demand points are captured by a solution when the features $f \in F$ do not occupy any selected planning units. Each demand point $i \in I$ for each $s \in S$ and $f \in F$ is assigned to an "imaginary" planning unit $j=J$ at $r=R$. The distance variables associated with this imaginary planning unit $d_{fsiJ}$ denote the loss of biological value associated with failing to secure a representative sample of feature $f$ in attribute space $s$. However, the $d$ variables are in distance units which effectively meaningless in this context. Thus these variables are calculated using a failure multiplier ($M$) and the maximum distance between the planning units and the demand points for $f \in F$, $s \in S$ (eqn 5).

@@

Conservation planning problems often involve several hundred or more planning units. It is currently not be feasible to solve problems this large when considering all possible failure scenarios (see Appendix S2). To reduce computational burden, users can set a smaller number to denote the maximum number of backup levels ($R$). For instance, when $R=3$ only 2 backup levels are considered in addition to the final backup level. Cui _et al._ [-@r16] found that $R=5$ yields comparable solutions to when $R$ is much larger. However, in most cases the decision maker will be limited to $R=1$ to obtain prioritizations in a feasible amount of time.

@@

\begin{align*}
R &= \parbox{25em}{number of backup levels} \tag*{eqn 6a} \\
%
M &= \parbox{25em}{failure multiplier} \tag*{eqn 6b}\\
%
\end{align*}

@@

\begin{align*}
Y_{fsijr} &= \begin{cases}
    1, & \parbox{25em}{if demand point $i$ for feature $f$ in space $s$ is conserved by planning unit $j$ at back-up level $r$. } \tag*{eqn 7a} \\
    0, & \parbox{25em}{otherwise} \\
  \end{cases} \\
%
P_{fsijr} &= \parbox{25em}{probability that demand point $i$ for feature $f$ in space $s$ will need to be conserved by planning unit $j$ at back-up level $r$} \tag*{eqn 7b}\\
%
\end{align*}

@@

The reliable formulation is non-linear. However, the non-linear components can be linearized. First---as discussed in the main text---the expression $X_j X_k$ in (eqn 3a) can be linearized using methods described by Beyer et al. [-@r426]. Second, the expression $P_{fsijr} Y_{fsijr}$ in (eqn 8a) can be linearized using techniques described by Sherali and Alameddine [-@r20] as implemented in Cui et al. [-@r16].

@@

We simulated conservation planning data sets containing different numbers of features and planning units (see Figure S2 for amounts). For a given number of planning units and features, a data set was simulated using the following steps. First, the planning units were simulated as squares that were arranged in a grid. Second, the planning unit costs were set to one. Third, the spatial distribution of each feature in the planning units were simulated using binary, spatially autocorrelated random fields [using the \texttt{RandomFields R} package; @r494; @r493]. Fourth, if a feature was found to occupy less than two planning units, the feature was set to occupy two randomly selected planning units. All occupied planning units were set to have a probability of occupancy of `r benchmark_parameters$occupancy_probability` for all features. Fifth, similar to the simulated data sets described in the main text, a geographic attribute space was generated for each feature. The features' demand points were set as the centroids of the planning units that they occupied, and were weighted by the units' probability of occupancy. After simulating the data sets, solutions were generated using the \texttt{raptr R} package and \texttt{Gurobi} (version 7.0.2).

We generated solutions for the conservation planning data sets using different formulations and parameters. Specifically, the solutions were generated using the unreliable and reliable formulation, and two different boundary length multipliers (`r paste(unique(benchmark_parameters$blm), collapse = " and ")`). The solutions were generated using amount-based targets of `r benchmark_parameters$amount_target*100` % and space-based targets of `r benchmark_parameters$space_target*100` %. To assess variation in solve times, we generated and solved `r benchmark_parameters$replicates` random data sets per combination of number of features and planning units. All problems were solved to within `r general_parameters$MIPGap * 100` % of optimality and were solved using a single thread on a system with 64 GB RAM and an Intel(R) Xeon (R) CPU (E5-2630 v3; 2.40 GHz).

@@

The benchmark analysis revealed that moderately-sized conservation problems can be solved in a relatively short period of time (Figure S2). Problems were solved much quicker when using the unreliable formulation than when using the reliable formulation. The solve times could be further reduced by using more threads. We were not able to determine how long bigger problems would take to solve using the reliable formulation because the system did not have enough RAM. Future research could involve implementing more advanced methods to solve the problems [eg. as described in @r16; @r15].

@@

# Figures
```{r, echo = FALSE, fig.height = 8, fig.width = 8.5, fig.cap = "Attribute spaces used in the first case-study. Each panel shows a the distribution of a solution in environmental space and how it samples the realized niche for a different species. The left column of panels shows the solution generated using amount targets. The right column of panels shows the solution generated using amount and space targets. Each column of panels corresponds to a different species. Hexagons show the distribution of demand points. The color of each hexagon denotes the weighted frequency of demand points inside it. Points denote the environmental conditions associated with planning units inside the species geographic range that were selected for preservation in a given solution."}
## prepare data
cs1_pu_niche_data <- plyr::ddply(cs1_spp_results,
  c("Species", "Prioritisation"),
  function(x) {
    # init
    curr_prioritisation <- cs1_prioritisations[[as.character(x$Prioritisation)]]
    curr_spp_pos <- match(x$Species[1], curr_prioritisation@data@species[[1]])
    curr_ids <- curr_prioritisation %>%
                slot("data") %>%
                slot("attribute.spaces") %>%
                `[[`(1) %>%
                slot("spaces") %>%
                `[[`(curr_spp_pos) %>%
                slot("planning.unit.points") %>%
                slot("ids")
    curr_sel_pos <- which(curr_ids %in% which(as.logical(
                          raptr::selections(curr_prioritisation))))
    # extract planning units
    curr_data <- curr_prioritisation %>%
                slot("data") %>%
                slot("attribute.spaces") %>%
                `[[`(1) %>%
                slot("spaces") %>%
                `[[`(curr_spp_pos) %>%
                slot("planning.unit.points") %>%
                slot("coords") %>%
                as.data.frame() %>%
                magrittr::set_names(c("Niche1", "Niche2")) %>%
                dplyr::mutate(Species = x[["Species"]][[1]],
                              Prioritisation = x[["Prioritisation"]][[1]],
                              Solution <- "discarded")
    curr_data$Solution[curr_sel_pos] <- "selected"
    # return data
    return(curr_data)
})
cs1_pu_niche_data %<>% dplyr::filter(Solution == "selected")

@@

```{r, echo = FALSE, message = FALSE, fig.height = 5.0, fig.width = 4.5, fig.cap = "Time required to solve different sized problems using different parameters and formulations. Points denote average times and bars show standard errors. Time is shown on a log$_{10}$ scale."}
# prepare data
plot_data <- benchmark_results %>%
  dplyr::mutate(blm = paste0("BLM = ", blm)) %>%
  dplyr::mutate(number_features = factor(number_features)) %>%
  dplyr::mutate(formulation = paste0(formulation, " formulation")) %>%
  dplyr::mutate(number_planning_units = factor(number_planning_units))
# make plot
ggplot2::ggplot(plot_data,
                ggplot2::aes(x = number_planning_units, y = time,
                             color = number_features)) +
  ggplot2::stat_summary(fun.y = mean, geom = "point",
                        position = ggplot2::position_dodge(0.95)) +
  ggplot2::stat_summary(fun.data = ggplot2::mean_se, geom = "errorbar",
                        position = ggplot2::position_dodge(0.95)) +
  ggplot2::facet_grid(blm ~ formulation ) +
  ggplot2::scale_y_continuous(trans = "log10") +
  ggplot2::xlab("Number of planning units") +
  ggplot2::ylab("Solve time (s)") +
  ggplot2::labs(color = "Number of features") +
  ggplot2::theme(
    axis.line.x = ggplot2::element_line(),
    axis.line.y = ggplot2::element_line(),
    strip.background = ggplot2::element_rect(fill = "grey20"),
    strip.text = ggplot2::element_text(color = "white"),
    legend.position = "bottom")
```

@@

```{r, echo = FALSE, results = "asis"}
# make table
terms <- c(
  "$A_j$", "area of planning unit $j$---note this can also be some other measure of conservation utility",
  "$B$", "boundary length modifier (BLM)",
  "$d_{fsij}$", "distance between demand point $i$ and planning unit $j$ for feature $f$ in attribute space $s$",
  "$e_{jk}$", "shared edge between planning unit $j$ and planning unit $k$. Note that when $j==k$ this used to parametrize exposed edges with no neighbours.",
  "$F$", "set of biodiversity features (indexed by $f$)",
  "$I_{fsi}$", "set of demand points (indexed by $i$) for a feature $f$ in attribute space $s$",
  "$J$", "set of planning units (indexed by $j$)",
  "$M$", "failure multiplier",
  "$P_{fsijr}$", "probability that demand point $i$ for feature $f$ in space $s$ will need to be conserved by planning unit $j$ at back-up level $r$",
  "$q_{fj}$", "probability of feature $f$ occupying planning unit $j$",
  "$R$", "number of backup levels (indexed by $r$)",
  "$S$", "set of attribute spaces (indexed by $S$)",
  "$T_f$", "amount target for feature $f$",
  "$X_j$", "binary decision variable controlling if a planning unit is selected for preservation (1) or discarded (0)",
  "$Y_{fsij}$", "binary decision variable indicating if planning unit $j$ is assigned to demand point $i$ for species $s$ in attribute space $s$ when determining the amount of the attribute space sampled by the selected planning units (unreliable version)",
  "$Y_{fsijr}$", "binary decision variable indicating if planning unit $j$ is assigned to demand point $i$ for species $s$ in attribute space $s$ at $r$-level $r$ when determining the amount of the attribute space sampled by the selected planning units (reliable version)",
  "$\\delta_{fsi}$", "the distance between each demand point $i$ and the centroid of the demand points $I$ for feature $f$ in attribute space $s$",
  "$\\lambda_{fsi}$", "set of weights for demand point $i$ for feature $f$ in attribute space $s$",
  "$\\tau_{fs}$", "space-based target for feature $f$ in attribute space $s$"
)
term_data <- data.frame(
  Symbol = terms[seq(1, length(terms), 2)],
  Description=terms[seq(2, length(terms), 2)]
)
term_data$Description <- paste0("\\parbox[t][][t]{15cm}{",
                                term_data$Description, "}")

# render table
curr_file <- tempfile(fileext = ".tex")
tmp <- Hmisc::latex(term_data,
                    file = curr_file,
                    digits = 2, rowname = NULL, first.hline.double = FALSE,
                    booktabs = TRUE, here = TRUE,
                    colheads = c("Symbol", "Description"),
                    col.just = c("c", "l"),
                    caption = paste0("Symbols and descriptions of terms used ",
                                     "in the formulation of the unreliable ",
                                     "representative and adequate ",
                                     "prioritization (URAP) problem and reliable version (RRAP)."))
curr_file %>%
  readLines() %>%
  `[`(-1) %>%
  paste0(collapse = "\n") %>%
  gsub(pattern = "toprule", replacement = "toprule[1pt]", fixed = TRUE) %>%
  gsub(pattern = "bottomrule", replacement = "bottomrule[1pt]",
       fixed = TRUE) %>%
  cat()
```
